\Wxxsir methods to estimate $\rr$ in the SIR model with applications to 2009 pandemic influenza

### 2018-08-23 paper/Version16/draft_v16.tex
Put in dependency for lag function

\title{\Wxxsir methods to estimate $\rr$ in the SIR model}
\author{ Shannon Gallagher$^{\dag}$, Andersen Chang$^{\ddag}$, and William F. Eddy$^{\dag}$ \\$\dag$ Department of Statistics and Data Science, Carnegie Mellon University\\ $\ddag$ Department of Statistics, Rice University}
\date{\today}
\maketitle

\begin{abstract}
For nearly a century, the initial reproduction number ($\rr$) has been used to summarize infectious diseases and to compare them to one another.  It is difficult to estimate $\rr$ due to assumptions on both how a disease transmits through a population and also due to differences in statistical estimation methods.  We describe nine methods used to estimate $\rr$ and provide a thorough simulation study of how these methods differ in the presence of disease parameters.  As real world motivation, we analyze the 2009 outbreak of the H1N1 pandemic influenza in the USA and compare the results from our nine methods to a previous study.  We discuss the most important aspects from our results which effect the estimation of $\rr$ and provide guidelines for estimating point estimates and confidence/credible intervals to create reliable, comparable estimates of $\rr$.
  \end{abstract}

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics'' \citep{Heesterbeek2002},  $\mathcal{R}_0$, the initial reproduction number (by convention pronounced ``R-naught'') is an elusive quantity for epidemic modelers to estimate.  As defined by \citet{anderson1992}, $\rr$ is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  In some ways, $\rr$ summarizes an entire outbreak of a disease; it is used to assess whether a disease outbreak will occur and its severity.  Additionally, it describes what percentage of the population needs to be vaccinated to avoid such an epidemic, roughly $1-\rr^{-1}$, tells us the final size of the total number of infected individuals, and describes the probability of which we will observe an outbreak under the same conditions \citep{anderson1992,britton2010}).  Despite a clear definition of $\rr$, epidemiologists have struggled to create a standard  estimator for $\rr$  \citep{hethcote2000}.  A major issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population \citep{diekmann2009}.

@@

\subsection{Smoothed Incidence to Prevalence Ratio (SIPR)}
We use the same method as above, IPR, but first smooth the epidemic curves $\left(\hat{X}(t), \hat{Y}(t) \right)$ (we use B-splines with $K=4$ degrees of freedom).  Then  $J(t) \approx -(\hat{X}(t+1) - \hat{X}(t))$, and the $\hat{\textnormal{IPR}}(t) = \frac{J(t)}{\hat{Y}(t)}$.  Then
\begin{align*}
\rr &= \hat{\textnormal{IPR}}(t) \cdot \frac{1}{\gamma}
\end{align*}
We use splines to smooth the curves because they  have proven to be good approximations of epidemic curves \citep{brooks2015}.  However, the researcher can use their method of choice to estimate $\hat{X}(t)$ and $\hat{Y}(t)$.  An advantage of this method is that it creates a less variable estimate than estimating IPR using only one point.  It has the same disadvantages as the usual IPR ratio in that it requires knowledge about $\gamma$.

@@

@@

Described by \cite{bettencourt2008} and summarized in \cite{obadia2012r0}, the sequential Bayes method is a Bayesian approach to an approximation of the classic SIR model.  The approximated SIR model assumes that the incidence at $t+1$, $J(t+1)$ has a Poisson distribution, with $\gamma$ as the  average inverse of the infectious period. In order to estimate $\rr$, we must have some idea about $\gamma$,
\begin{align*}
J(t+1) | \rr  \sim \textnormal{Poisson}( J(t) \exp \left \{  \gamma (\rr-1)\right \})
\end{align*}
Then, the posterior distribution of $\rr$ given the previous days' incidence is
\begin{align*}
  P(\rr | J_0, \dots, J_{t+1}) = \frac{P(J_{t+1} | \rr, J_0, \dots, J_t)P(\rr| J_0, \dots, J_t)}{P(J_0, \dots, J_{t+1})}.
\end{align*}
This method is sequential in that the prior distribution for $\rr$ comes from the previous day.  The initial prior for $\rr$ is assumed to be flat.  This method results in a posterior distribution from which credible intervals may be obtained.  This method assumes, initial growth in incidence to be exponential, and homogeneous mixing of populations as with any compartment model.  The advantages of this method are that of the ability to obtain an entire posterior distribution, whereas many other methods are difficult to even find an estimate of the variance.  Disadvantages include strict assumptions about the distributions and computational time required to estimate the relevant parameters.

@@

We compare the performance of the SIR models on simulated data from both the SIR as well as other models. In all of these simulations, we first generate data from the models under known functions ($f$, $g$), then we add noise drawn from known distributions to each time point $(\epsilon_{X,t}, \epsilon_{Y,t})$,
\begin{align}\label{eq:sim-models}
  \hat{X}(t) &= f(t) + \epsilon_{X,t} \\
  \hat{Y}(t) &= g(t) + \epsilon_{Y,t} \nonumber\\
  \hat{Z}(t) &= N - \hat{X}(t) - \hat{Y}(t)\nonumber 
\end{align}
We adjust the recovered compartment $Z$ so the total population is constant. 
We simulate with two different types of error distributions, Gaussian (Norm) and autoregressive (AR),

\noindent \textbf{Gaussian (Norm) ($\hat{X}_{G}$, $\hat{Y}_{G}$, $\hat{Z}_{G}$)}
\begin{align*}
  f(t) &= X_{XYZ}(t) \\
  g(t) &= Y_{XYZ}(t) \\
  \epsilon_{X,t} &\sim N(0, \sigma_X^2) \\
  \epsilon_{Y,t} &\sim N(0, \sigma_Y^2)
\end{align*}
\textbf{Autoregressive(1) (AR) ($\hat{X}_{AR}$, $\hat{Y}_{AR}$, $\hat{Z}_{AR}$)}
\begin{align*}
  f(t) &= \rho \hat{X}(t-1) \\
  g(t) &= \rho \hat{Y}(t-1) \\
  \epsilon_{X,t} &\sim N(0, \sigma_X^2) \\
  \epsilon_{Y,t} &\sim N(0, \sigma_Y^2).\\
\end{align*}
For each of the Gaussian and AR datasets, we form an additional dataset which enforces the monotonicity of X and Z via the Pool Adjacent Violators Algorithm (PAVA), described by \cite{friedman1984}).  Here, we adjust $Y$ so the constant population is maintained.

\noindent \textbf{Gaussian-PAVA (Norm-M) ($\hat{X}_{GP}$, $\hat{Y}_{GP}$, $\hat{Z}_{GP}$)}
\begin{align*}
 X_{GP}(t) &= \textnormal{PAVA}(X_G(t)) \\
  Y_{GP}(t) &= N - X_{GP}(t) - Z_{GP}(t) \\
  Z_{GP}(t) &= \textnormal{PAVA}(Z_G(t)) \\
\end{align*}
\textbf{Autoregressive-PAVA (AR-M) ($\hat{X}_{ARP}$, $\hat{Y}_{ARP}$, $\hat{Z}_{ARP}$)}
\begin{align*}
  X_{ARP}(t) &= \textnormal{PAVA}(X_{AR}(t)) \\
  Y_{ARP}(t) &= N - X_{ARP}(t) - Z_{ARP}(t) \\
  Z_{ARP}(t) &= \textnormal{PAVA}(Z_{AR}(t)) 
\end{align*}
This gives us a total of four error structures for each set of conditions  (1) Gaussian (Norm), (2) Gaussian monotone (Norm-M),  (3) autoregressive (AR), and (4) autoregressive monotone (AR-M) for a given set of parameters $\left (\beta, \gamma, T, X(0), Y(0), Z(0), \sigma_X, \sigma_Y, N \right )$, the infection rate; recovery rate; total number of time points; initial conditions for number of susceptibles, infectious, and recovered individuals; the standard deviations for both the susceptible and infectious compartments; and the total population, respectively.

@@

      \end{center}
      The baseline data set is generated by adding noise to the susceptible and infectious curves from  deterministic SIR, shown in Figure \ref{fig:baseline-data}.  The recovered compartment is then adjusted to maintain a constant population value.  In this scenario, we see that the infection begins slowly but over the course of a year infects nearly 75\% of the population.
      \begin{figure}
        \centering
        \includegraphics[width=.8\textwidth]{images/baseline_sir.pdf}
        \caption{SIR (XYZ) curves for the deterministic model with $N=100,000$, $(\beta, \gamma) = (0.06, 0.03)$, $(X(0), Y(0), Z(0))= (99,950, 50,0)$  The baseline data is generated by using this model and adding noise to the $X$ and $Y$ compartments and adjusting $Z$ at each time step to maintain a constant population.}\label{fig:baseline-data}
        \end{figure}
      This baseline data set is chosen due to its moderate $\rr$ size of 2, which indicates an outbreak of an epidemic similar to HIV, ZIK-V, EVD, and H1N1 pandemic influenza (see Table \ref{tab:r0-real-ex}).  The infection rate and recovery rate ($\beta, \gamma$) are relatively small.  The duration of the disease, under the true model, spans nearly a whole year (see Figure \ref{fig:baseline-data}) and so we use a year's worth of daily observations of ($\hat{X},\hat{Y}, \hat{Z}$) values.  Initially, 50 (0.05\%) infectious individuals are introduced to the population of 100,000 individuals (approximately the average population size of a US county in 2015).  The standard deviation of the $\hat{X}$ values is 0.1\% of the magnitude of the initial number of susceptible individuals, and the standard deviation of the $\hat{Y}$ values is 10\% of the magnitude of the initial number of infectious individuals, although in this setting the number of infectious individuals increases rapidly.  We use Gaussian error as our baseline error because of its well known statistical properties so we can more easily compare methods to one another.  Later on, we explore the autoregressive and PAVA simulated data, which may better simulate data from  a real epidemic.

@@

  Baseline        & SIR      &   Norm             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\ \midrule

@@

%  \textcolor{red}{Norm-M}      & SIR      &   Norm-M ($\rho=.7)$             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                            \\ \midrule

@@

  InfRec1       & SIR      &   Norm             & (60000, .001)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)\\
  InfRec2      & SIR      &   Norm             & (1500, .001)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\
  InfRec3      & SIR      &   Norm             & (1000, .001)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\
InfRec4      & SIR      &   Norm             & (06, .24)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\ \midrule
  Time1 & SIR &  Norm   & (.06, .03)                          &     200       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\
  Time2 & SIR &  Norm   & (.06, .03)                          &     100       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\
  Time3 & SIR &  Norm   & (.06, .03)                          &     50       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\
  Time4 & SIR &  Norm   & (.06, .03)                          &     20       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\ \midrule
  Initial1    & SIR      &   Norm             & (.06, .03)                          &     365       &  (99999, 1)                    &  $10^5$          &(100, 5)                                \\
  Initial2    & SIR      &   Norm             & (.06, .03)                          &     365       &  (99990, 10)                    &  $10^5$          &(100, 5)                                \\
  Initial3    & SIR      &   Norm             & (.06, .03)                          &     365       &  (99900, 100)                    &  $10^5$          &(100, 5)                                \\
  Initial4    & SIR      &   Norm             & (.06, .03)                          &     365       &  (99000, 1000)                    &  $10^5$          &(100, 5)                                \\ \midrule
PopSize1      & SIR      &   Norm             & (.06, .03)                          &     365       &  (999500, 500)                    &  $10^6$          &(100, 5)                                \\
PopSize2      & SIR      &   Norm             & (.06, .03)                          &     365       &  (9995, 5)                    &  $10^4$          &(100, 5)                                \\
PopSize3        & SIR      &   Norm             & (.06, .03)                          &     365       &  (990, 10)                    &  $10^3$          &(100, 5)                                \\
PopSize4       & SIR      &   Norm             & (.06, .03)                          &     365       &  (99, 1)                    &  $10^2$          &(100, 5)                                \\

@@

  Var1      & SIR      &   Norm             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(50, 2)                                \\
  Var2      & SIR      &   Norm             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(500, 20)                                \\
  Var3      & SIR      &   Norm             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(2500, 1000)                                \\
  Var4      & SIR      &   Norm             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(10000, 500)                                \\ \midrule
 Poly1     & Linear      &   Norm             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                            \\
Poly2    & Quartic      &   Norm             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                            \\ 
   LSIR     & LSIR      &   Norm             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                            \\ 

@@

\caption{Table of simulated data sets for the ``Norm'' error.  We also repeat the tables with the other three error structures: AR, Norm-M, and AR-M.  We simulate $4 \times 24=96$ total data sets with the given parameters.}

@@

Table \ref{tab:se-methods} shows the ways that standard errors are acquired for the different methods. The IPR, and SIPRknowledge of the length of the average infectious period $\gamma^{-1}$. For the results below, we assume the best case scenario that the true average infectious period is known for the data set.

\begin{table}[H]
	\centering
	\begin{tabular}{@{}ll@{}}
		\toprule
		\textbf{Model} & \textbf{Default Standard Error Method} \\ \midrule
		Ratio Estimator (RE) & Delta (with inverse Hessian)\\
		Reparameterized Ratio Estimator (rRE) & Delta (with inverse Hessian) \\
		Linear Model Approximation (LMA) & Delta (linear regression SE) \\
		LMA, All Time Points  (LMAT)& Delta (linear regression SE)\\
		Incidence-Prevalence Ratio (IPR) & Block bootstrap (Jackknife resampling) \\
		Smoothed IPR (SIPR) & Block bootstrap (Jackknife resampling) \\
		Log-Linear (LL) & Delta (linear regression SE) \\
		Markov Chain (MC) & Delta (GLM regression SE) \\
		Sequential Bayes (SB) & Posterior\\
		% Exponential Growth (EG) & Delta method with linear regression standard error \\
		% IDEA & Delta method with linear regression standard error \\
		\bottomrule
	\end{tabular}
	\caption{Table of methods for estimating standard errors for each of our $\rr$ estimates.}
	\label{tab:se-methods}
\end{table}

@@

In particular, MC and SB almost never capture the true estimate and typically underestimate the true value of $\rr$.  This leads us to believe that these models are inadequate in capturing the effects of a disease following the SIR model.  Both models rely on parametric assumptions of simple underlying statistical distributions for the course of infection (binomial and Poisson, respectively) with time-invariant transmission parameters ($\alpha$ and $\lambda$).  These are good examples of why we need to consider other reasons for using a model other than the math resulting in a simple estimate.  That is not to say that variants of these models are unviable for modelling $\rr$, but they do not make for good defaults.  It is also an issue that these two methods consistently underestimate $\rr$, which could possibly lead to an unchecked epidemic.

@@

%% robust to different error assumptions
 In Figure \ref{fig:err-rob} we plot the standardized residuals of $\rr$ focusing on the three best methods: RE, rRE, and LL.  Reassuringly, we find that $\hat{\rr}$ estimates from RE and rRE seem to be generally robust to the error structure.  ((What's up with varying $\rr$ values??))

@@

@@

>>>>

\title{\Wxxsir methods to estimate $\rr$ in the SIR model with applications to 2009 pandemic influenza}
\author{ Shannon Gallagher$^{\dag}$, Andersen Chang$^{\ddag}$, and William F. Eddy$^{\dag}$ \\$\dag$ Department of Statistics and Data Science, Carnegie Mellon University\\ $\ddag$ Department of Statistics, Rice University}
\date{\today}
\maketitle

\begin{abstract}
For nearly a century, the initial reproduction number ($\rr$) has been used to summarize infectious diseases and to compare them to one another.  Difficulties in estimating $\rr$ arise from on both how a disease transmits through a population as well as from differences in statistical estimation methods.  We describe \wxxsir methods used to estimate $\rr$ and provide a thorough simulation study of how these methods differ in the presence of disease parameters.  As real world motivation, we analyze the 2009 outbreak of the H1N1 pandemic influenza in the USA and compare the results from our nine methods to a previous study.  We discuss the most important aspects from our results which effect the estimation of $\rr$ and provide guidelines for estimating point estimates and confidence/credible intervals to create reliable, comparable estimates of $\rr$.
  \end{abstract}

\section{Introduction}\label{sec:intro}
What has been called ``arguably the most important quantity in the study of epidemics,''  $\mathcal{R}_0$, the initial reproduction number (by convention pronounced ``R-naught'') is an elusive quantity for epidemic modelers to estimate \citep{Heesterbeek2002}.  As defined by \citet{anderson1992}, $\rr$ is the ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible.''  In some ways, $\rr$ summarizes an entire outbreak of a disease; it is used to assess whether a disease outbreak will occur and its severity.  Additionally, it describes what percentage of the population needs to be vaccinated to avoid such an epidemic, roughly $1-\rr^{-1}$, tells us the final size of the total number of infected individuals, and describes the probability of which we will observe an outbreak under the same conditions \citep{anderson1992,britton2010}).  Despite a clear definition of $\rr$, epidemiologists have struggled to create a standard  estimator for $\rr$  \citep{hethcote2000}.  A major issue in estimating $\rr$ is that the quantity is a \textit{property of the model}, meaning that $\rr$ is dependent not only on the usual noise that comes with statistical modeling but also on a variety of assumptions on how researchers assume a disease is transmitted through a population \citep{diekmann2009,brown2016}.

@@

@@

%%%%%%%%%%%%55
\subsection{Likelihood Based Estimation (LBE)}\label{sec:likelihood-based}
A common way used to estimate epidemiological quantities is through likelihood based estimation.  We saw an example of this in the previous section with MC estimation.  However, as we will see in the results, this particular model does not seem to fit the data well.  As a result, many researchers specify different models for the generation of SIR data; estimate the likelihood through direct calculation, a method such as iterated filtering used in ``plug-and-play'' and derive point estimates and CIs from this likelihood estimate  \citep{abbey1952,forsberg2008,king2009,bhadra2012}.  
%https://kingaa.github.io/sbied/stochsim/stochsim.html

Here, we examine one model that has been used to model the (discrete) SIR: A binomial chain with probabilities based on the deterministic SIR model.

The discrete XYZ model is as follows, where we have replaced $d$ with $\Delta$,
\begin{align*}
\systeme{\frac{\Delta X}{\Delta t} = -\frac{\beta Y X}{N}, \frac{\Delta Y}{\Delta t} = \frac{\beta Y X}{N} - \gamma Y, \frac{\Delta Z}{\Delta t} = \gamma Y}. 
\end{align*}

Denote $\hat{X}(t), \hat{Y}(t)$, and $\hat{Z}(t)$ to be the observed number in each compartment and $X(t), Y(t)$, and $Z(t)$ to be the true, underlying model at time $t$.   The stochastic model is given by
\begin{align*}
  \hat{X}(t+1) &= \hat{X}(t) - x_t\\
  \hat{Y}(t+1) & = N - \hat{X}(t+1) - \hat{Z}(t+1), \nonumber \\
  \hat{Z}(t+1) &= \hat{Z}(t) + z_t \nonumber
\end{align*}
with $\hat{X}(0)$, $\hat{Y}(0)$, and $\hat{Z}(0)$ known. The random terms in Equation \eqref{eq:sir-noise} are random variables.  Here we choose them to have the following distributions,
\begin{align*}
  x_{t} &\sim \text{Binomial}\left (\hat{X}(t), \frac{\beta Y(t)}{N} \right ) \\
  z_{t} &\sim \text{Binomial}\left (\hat{Y}(t), \gamma \right ). \nonumber
\end{align*}
Then the likelihood is
\begin{align*}
  \mathcal{L}\left (\beta, \gamma ; \textnormal{Data} \right ) &=
                                                  \prod_{t=0}^{T-1} \left [\left ( \begin{array}{c}
                                                            \hat{X}(t) \\
                                                            x_t
                                                            \end{array}
  \right ) \left (\beta - \frac{\beta\left (X(t) +Z(t) \right )}{N} \right )^{x_t} \left (1 - \beta +  \frac{\beta\left (X(t) + Z(t) \right )}{N} \right )^{\hat{X}(t) - x_t} \cdot \right .\\
  & \left . \left ( \begin{array}{c}
              N -\left ( \hat{X}(t) + \hat{Z}(t)\right ) \\
              z_t
            \end{array}
  \right ) \left (\gamma \right )^{z_t} \left ( 1 - \gamma\right )^{N -\left ( \hat{X}(t) + \hat{Z}(t)\right ) - z_t} \right ].
\end{align*}
The estimate for $\rr$ is then
\begin{align}\label{eq:lbe-r0}
  (\hat{\beta}, \hat{\gamma}) &= \arg \max_{\beta, \gamma} \mathcal{L}\left ( \beta, \gamma; \textnormal{Data} \right )\nonumber\\
  \rr &= \frac{\hat{\beta}}{\hat{\gamma}}
  \end{align}
  The two main differences between this model and MC are  that 1) LBE looks at the joint likelihood of the XYZ model, whereas MC only explicitly models the number of susceptibles and 2) the difference in the structure of the transition probability.  We use the delta method in conjunction with $\hat{\beta}$ and $\hat{\gamma}$ estimations from the Nelder-Mead optimization of the likelihood to estimate the stantard error of $\hat{\rr}$.
  The advantages of using this model are that our modelling assumptions are explicit and that the modelling process seeems to be reasonable.  Disadvantages include maximizing the liklihood may be more time consuming than other methods due to the need to approximate $X(t)$ and $Z(t)$ for many different values of $\beta$ and $\gamma$ and that if we are wrong about the generating process of the XYZ curves, then this model may be a very poor fit.

@@

Described by \cite{bettencourt2008} and summarized in \cite{obadia2012r0}, the sequential Bayes method is a Bayesian approach to an approximation of the classic SYR model.  The approximated SIR model assumes that the incidence at $t+1$, $J(t+1)$ has a Poisson distribution, with $\gamma$ as the  average inverse of the infectious period. In order to estimate $\rr$, we must have some idea about $\gamma$,
\begin{align*}
J(t+1) | \rr  \sim \textnormal{Poisson}( J(t) \exp \left \{  \gamma (\rr-1)\right \})
\end{align*}
Then, the posterior distribution of $\rr$ given the previous days' incidence is
\begin{align*}
  P(\rr | J_0, \dots, J_{t+1}) = \frac{P(J_{t+1} | \rr, J_0, \dots, J_t)P(\rr| J_0, \dots, J_t)}{P(J_0, \dots, J_{t+1})}.
\end{align*}
This method is sequential in that the prior distribution for $\rr$ comes from the previous day.  The initial prior for $\rr$ is assumed to be flat.  This method results in a posterior distribution from which credible intervals may be obtained.  This method assumes, initial growth in incidence to be exponential, and homogeneous mixing of populations as with any compartment model.  The advantages of this method are that of the ability to obtain an entire posterior distribution, whereas many other methods are difficult to even find an estimate of the variance.  Disadvantages include strict assumptions about the distributions and computational time required to estimate the relevant parameters.

@@

We compare the performance of the SIR models on simulated data from both the SIR as well as other models. In all of these simulations, we first generate data from the models under known functions ($f$, $g$), then we add noise drawn from known distributions to each time point $(\epsilon_{X,t}, \epsilon_{Y,t})$,
\begin{align}\label{eq:sim-models}
  \hat{X}(t) &= f(t) + \epsilon_{X,t} \\
  \hat{Y}(t) &= g(t) + \epsilon_{Y,t} \nonumber\\
  \hat{Z}(t) &= N - \hat{X}(t) - \hat{Y}(t)\nonumber 
\end{align}
We adjust the recovered compartment $Z$ so the total population is constant. 
We simulate with two different types of error distributions, Gaussian (G) and autoregressive (AR),

\noindent \textbf{Gaussian (G) ($\hat{X}_{G}$, $\hat{Y}_{G}$, $\hat{Z}_{G}$)}
\begin{align*}
  f(t) &= X_{XYZ}(t) \\
  g(t) &= Y_{XYZ}(t) \\
  \epsilon_{X,t} &\sim N(0, \sigma_X^2) \\
  \epsilon_{Y,t} &\sim N(0, \sigma_Y^2)
\end{align*}
\textbf{Autoregressive(1) (AR) ($\hat{X}_{AR}$, $\hat{Y}_{AR}$, $\hat{Z}_{AR}$)}
\begin{align*}
  f(t) &= \rho \hat{X}(t-1) \\
  g(t) &= \rho \hat{Y}(t-1) \\
  \epsilon_{X,t} &\sim N(0, \sigma_X^2) \\
  \epsilon_{Y,t} &\sim N(0, \sigma_Y^2).\\
\end{align*}
For each of the Gaussian and AR datasets, we form an additional dataset which enforces the monotonicity of X and Z via the Pool Adjacent Violators Algorithm (PAVA), described by \cite{friedman1984}).  Here, we adjust $Y$ so the constant population is maintained, since X and Z are both monotone functions.

\noindent \textbf{Gaussian-Monotone-PAVA (G-M) ($\hat{X}_{GP}$, $\hat{Y}_{GP}$, $\hat{Z}_{GP}$)}
\begin{align*}
 X_{GP}(t) &= \textnormal{PAVA}(X_G(t)) \\
  Y_{GP}(t) &= N - X_{GP}(t) - Z_{GP}(t) \\
  Z_{GP}(t) &= \textnormal{PAVA}(Z_G(t)) \\
\end{align*}
\textbf{Autoregressive-Monotone-PAVA (AR-M) ($\hat{X}_{ARP}$, $\hat{Y}_{ARP}$, $\hat{Z}_{ARP}$)}
\begin{align*}
  X_{ARP}(t) &= \textnormal{PAVA}(X_{AR}(t)) \\
  Y_{ARP}(t) &= N - X_{ARP}(t) - Z_{ARP}(t) \\
  Z_{ARP}(t) &= \textnormal{PAVA}(Z_{AR}(t)) 
\end{align*}
This gives us a total of four error structures for each set of conditions  (1) Gaussian (G), (2) Gaussian monotone - PAVA (G-M),  (3) autoregressive (AR), and (4) autoregressive monotone - PAVA (AR-M) for a given set of parameters $\left (\beta, \gamma, T, X(0), Y(0), Z(0), \sigma_X, \sigma_Y, N \right )$, the infection rate; recovery rate; total number of time points; initial conditions for number of susceptibles, infectious, and recovered individuals; the standard deviations for both the susceptible and infectious compartments; and the total population, respectively.

@@

      \end{center}
      The baseline data set is generated by adding noise to the susceptible and infectious curves from deterministic SIR paths, which are shown in Figure \ref{fig:baseline-data}.  The recovered compartment is then adjusted to maintain a constant population value.  In this scenario, we see that the infection begins slowly but over the course of a year infects nearly 75\% of the population.
      \begin{figure}
        \centering
        \includegraphics[width=.8\textwidth]{images/baseline_sir.pdf}
        \caption{SIR (XYZ) curves for the deterministic model with $N=100,000$, $(\beta, \gamma) = (0.06, 0.03)$, $(X(0), Y(0), Z(0))= (99950, 50,0)$  The baseline data is generated by using this model and adding noise to the $X$ and $Y$ compartments and adjusting $Z$ at each time step to maintain a constant population.}\label{fig:baseline-data}
        \end{figure}
      This baseline data set is chosen due to its moderate $\rr$ size of 2, which indicates an outbreak of an epidemic similar to HIV, ZIK-V, EVD, and H1N1 pandemic influenza (see Table \ref{tab:r0-real-ex}).  The infection rate and recovery rate ($\beta, \gamma$) are relatively small.  The duration of the disease, under the true model, spans nearly a whole year (see Figure \ref{fig:baseline-data}) and so we use a year's worth of daily observations of ($\hat{X},\hat{Y}, \hat{Z}$) values.  Initially, 50 (0.05\%) infectious individuals are introduced to the population of 100,000 individuals (approximately the average population size of a US county in 2015).  The standard deviation of the $\hat{X}$ values is 0.1\% of the magnitude of the initial number of susceptible individuals, and the standard deviation of the $\hat{Y}$ values is 10\% of the magnitude of the initial number of infectious individuals, although in this setting, the number of infectious individuals increases rapidly.  We use Gaussian error as our baseline error because of its well known statistical properties so we can more easily compare methods to one another.  Later on, we explore the autoregressive and PAVA simulated data, which may better simulate data from  a real epidemic.

@@

  Baseline        & SIR      &   G             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\ \midrule

@@

%  \textcolor{red}{G-M}      & SIR      &   G-M ($\rho=.7)$             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                            \\ \midrule

@@

  InfRec1       & SIR      &   G             & (60000, .001)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)\\
  InfRec2      & SIR      &   G             & (1500, .001)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\
  InfRec3      & SIR      &   G             & (1000, .001)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\
InfRec4      & SIR      &   G             & (06, .24)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\ \midrule
  Time1 & SIR &  G   & (.06, .03)                          &     200       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\
  Time2 & SIR &  G   & (.06, .03)                          &     100       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\
  Time3 & SIR &  G   & (.06, .03)                          &     50       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\
  Time4 & SIR &  G   & (.06, .03)                          &     20       &  (99950, 50)                    &  $10^5$          &(100, 5)                                \\ \midrule
  Initial1    & SIR      &   G             & (.06, .03)                          &     365       &  (99999, 1)                    &  $10^5$          &(100, 5)                                \\
  Initial2    & SIR      &   G             & (.06, .03)                          &     365       &  (99990, 10)                    &  $10^5$          &(100, 5)                                \\
  Initial3    & SIR      &   G             & (.06, .03)                          &     365       &  (99900, 100)                    &  $10^5$          &(100, 5)                                \\
  Initial4    & SIR      &   G             & (.06, .03)                          &     365       &  (99000, 1000)                    &  $10^5$          &(100, 5)                                \\ \midrule
PopSize1      & SIR      &   G             & (.06, .03)                          &     365       &  (999500, 500)                    &  $10^6$          &(100, 5)                                \\
PopSize2      & SIR      &   G             & (.06, .03)                          &     365       &  (9995, 5)                    &  $10^4$          &(100, 5)                                \\
PopSize3        & SIR      &   G             & (.06, .03)                          &     365       &  (990, 10)                    &  $10^3$          &(100, 5)                                \\
PopSize4       & SIR      &   G             & (.06, .03)                          &     365       &  (99, 1)                    &  $10^2$          &(100, 5)                                \\

@@

  Var1      & SIR      &   G             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(50, 2)                                \\
  Var2      & SIR      &   G             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(500, 20)                                \\
  Var3      & SIR      &   G             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(2500, 1000)                                \\
  Var4      & SIR      &   G             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(10000, 500)                                \\ \midrule
 Poly1     & Linear      &   G             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                            \\
Poly2    & Quartic      &   G             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                            \\ 
   LSIR     & LSIR      &   G             & (.06, .03)                          &     365       &  (99950, 50)                    &  $10^5$          &(100, 5)                            \\ 

@@

\caption{Table of simulated data sets for the ``G'' error.  We also repeat the tables with the other three error structures: AR, G-M, and AR-M.  We simulate $4 \times 24=96$ total data sets with the given parameters.}

@@

Table \ref{tab:se-methods} shows the ways that standard errors are acquired for the different methods. The method IPR requires knowledge of the length of the average infectious period $\gamma^{-1}$ in order to fit the model. For the results below, we assume the best case scenario that the true average infectious period is known for the data set.

\begin{table}[H]
	\centering
	\begin{tabular}{@{}ll@{}}
		\toprule
		\textbf{Model} & \textbf{Default Standard Error Method} \\ \midrule
		Ratio Estimator (RE) & Delta (with inverse Hessian)\\
		Reparameterized Ratio Estimator (rRE) & Delta (with inverse Hessian) \\
		Linear Model Approximation (LMA) & Delta (linear regression SE) \\
		LMA, All Time Points  (LMAT)& Delta (linear regression SE)\\
		Incidence-Prevalence Ratio (IPR) & Block bootstrap (Jackknife resampling) \\
		Log-Linear (LL) & Delta (linear regression SE) \\
          Markov Chain (MC) & Delta (GLM regression SE) \\
          Likelihood Based Estimation (LBE) & Delta (with inverse Hessian)\\
		Sequential Bayes (SB) & Posterior\\
		% Exponential Growth (EG) & Delta method with linear regression standard error \\
		% IDEA & Delta method with linear regression standard error \\
		\bottomrule
	\end{tabular}
	\caption{Table of methods for estimating standard errors for each of our $\rr$ estimates.}
	\label{tab:se-methods}
\end{table}

@@

In particular, MC and SB almost never capture the true estimate and typically underestimate the true value of $\rr$.  This leads us to believe that these models are inadequate in capturing the effects of a disease following the SIR model.  Both models rely on parametric assumptions of simple underlying statistical distributions for the course of infection (Binomial and Poisson, respectively) with time-invariant transmission parameters ($\alpha$ and $\lambda$).  These are good examples of why we need to consider other reasons for using a model other than the math resulting in a simple estimate.  That is not to say that variants of these models are unviable for modelling $\rr$, but they do not make for good defaults.  It is also an issue that these two methods consistently underestimate $\rr$, which could possibly lead to an unchecked epidemic.

@@

%% robust to different error assumptions
 In Figure \ref{fig:err-rob} we plot the standardized residuals of $\rr$ focusing on the three best methods: RE, rRE, and LL.  Reassuringly, we find that $\hat{\rr}$ estimates from RE and rRE seem to be generally robust to the error structure.  \com{((What's up with varying $\rr$ values??))}

@@

\com{Table and graph do not agree}
\begin{table}[H]

@@

\com{Results TBD}
